Etox-Base: A tool to derive exposure endpoints from chemical test results

### 2019-02-08 article/article.tex

this data will be useful for

@@

plan:
- occurring in vast numbers
- pathways
- ecosystem services 
- biodiversity loss
- Übergang abschätzung von chemikalien

@@

The EPA releases \epa{} on a quarterly basis through their website. The software downloads each new version and rebuilds it locally in a PostgreSQL data base. A processing pipeline of several R \citep{r_core_team_r_2017} scripts prepares the data for the tool. The tool itself is a R function that is executed by a user accessing the R shiny \citep{chang_shiny_2018} application. In the following the steps for processing the data are explained in more detail (PATH-TO-OVERVIEW) as well as the scripts that have been used for it (REFER-TO-SCRIPTS-TABLE).

@@

The data is downloaded (bd\_epa\_download.R) and subsequently built into a PostgreSQL data base (bd\_epa\_postgres.R). Thereafter the data is imported into R for pre-processing (da\_epa1.R, da\_epa2.R, da\_epa3.R), which includes validity checks, removal of special characters, unit harmonizations and selection of relevant variables. At the end the data is stored locally (epa3.rds, epa2\_taxa.rds, epa2\_chem.rds). CAS numbers, unique identifiers and taxon names are then used to query other data bases for additional information on chemicals and biota respectively. Chemical information is retrieved from the Pubchem data base \citep{CITE_PUBCHEM} (qu\_pc.R), from the Compendium of Pesticide Common Names \citep{CITE_AW} (qu\_aw.R), from the Physprop data base \citep{CITE_PHYSPROP} (qu\_pp.R), from EUROSTAT and the Chemspider data base \citep{CITE-CHEMSPIDER}. These queries mostly rely on the webchem R-package \citep{szocs_webchem_2015-1}. For biota, habitat and occurrence information is queried from the World Register of Marine Species (WORMS) \citep{WORMS} and from the Global Biodiversity Information Facility (GBIF) \citep{CITE_RGBIF} using the rgbif R-package. The acquired information is then merged (re\_merge.R), variables are combined and created (re\_combine.R), checked (re\_checks\_internal.R) and finally compiled in one final table (re\_final.R) This table is then used in the application to perform search queries on. For a detailed overview of what information is collected see table \ref{table:processing-scripts}.

\subsection*{The application}
The application is accessible through \etoxbase{} and was built by using the shiny web application framework \citep{chang_shiny_2018} in R. Therein the user can choose to filter the data by several parameters such as chemical class, organism habitat, test duration and test endpoints amongst others. A detailed list for possible filter parameters can be found in \ref{table:meta}.

In order to be able to calculate TUs and SSDs, derivation of single endpoint values for specific organism groups and a chemicals of the partly extensive test data, the \etoxbase{} aggregates the test results according to the chosen filters in two steps. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon plus the duration and secondly only by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the mean or the geometric mean as aggregate. Besides the aggregation some cleaning can also be performed. The \etoxbase{} allows to exclude test results exhibiting concentrations that are higher than the actual water solubility of the respective chemical at \ang{20} C. In any case the user can choose to exclude outliers. The outliers are selected as values that exceed lower (0.25) and the upper (0.75) quartile by 1.5 times the inter-quartile range. All the steps described in this section are executed at each click in the app which runs the function fun\_ec50filter\_aggregation.R (cf. table \ref{table:processing-scripts}). \citep{dowle_data.table_2018}.

\subsection*{Future of the project}
As new versions of the \epa{} is published on a quarterly basis the data acquisition process is automated and new versions of the data and the \app{} are planed to be released regularly.

\begin{figure}
    \includestandalone[width=\textwidth, scale=0.1]{tikz/pipeline_organigram}
    \caption{Organigram}
    \label{fig:organigram}
\end{figure}

@@

The code for processing the data is stored in a Github reopsitory (\git{}). All processing scripts were written in R version 3.4.4. PostgreSQL 9.5 has been used to build the data base. Hence R together with its packages and PostgreSQL suffice to rebuild the process. The \app{} itself is stored in a second Github repostitory (\gitapp{}).

@@

\begin{figure}
    \includestandalone[scale=0.5]{tikz/tree_source}
    \caption{Processing pipeline}
    \label{fig:organigram}
\end{figure}

@@

\section*{Usage Notes}
%%Brief instructions that may help other researchers reuse these dataset.
%%This is an optional section, but strongly encouraged when helpful
%%to readers. This may include discussion of software packages that
%%are suitable for analyzing the assay data files, suggested downstream
%%processing steps (e.g. normalization, etc.), or tips for integrating
%%or comparing this with other datasets. If needed, authors are encouraged
%%to provide code, programs, or data processing workflows when they may help 
%%others analyse the data. We encourage authors to archive related code in 
%%a DOI-issuing archive when possible, but code may also be supplied as 
%%supplementary information files. 
%%
%%For studies involving privacy or safety controls on public access
%%to the data, this section should describe in detail these controls,
%%including how authors can apply to access the data, and what criteria
%%will be used to determine who may access the data, and any limitations
%%on data use.
This data set is designed to support ecotoxicologists in assessing the risk of chemicals on the environment. As there is more and more ecotoxicological test data available, it becomes fundamental to provide such information in reasonable formats. This means, that it should be accessible for humans as well as computers. Up to now ecotoxicoloists rely on effect values of single publications or data bases such as the Pesticide Property Data Base (PPDB), having certain limitations such as encompassing only a few substance groups. The \epa{} tries to fill these gaps by collecting in peer-review articles published ecotoxicological test results. However due to lacking test information and un-harmonized data sets it is not always easily accessible. The \app{} tries to mitigate this gap. An ecotoxicologist can now access the \app{} and retrieve cleaned, harmonized and aggregated test data according to his or her needs.

@@

@@

write here that it is meant as an addition to PPDB

@@

other approaches: https://envirotoxdatabase.org/ - performs no aggregation, not open source

@@

\begin{table}
%\begin{sidewaystable}
    \csvautotabular[respect underscore=true]{data/scripts.csv}
    \caption{R Scripts to query and process data obtained from various resources. The script tree, i.e. the order which script is processed can be found in}
    \label{table:processing-scripts};
%\end{sidewaystable}
\end{table}

\csvreader[tabular=|l|l|,
    table head=\hline & category & description\\\hline,
    late after line=\\\hline]
{data/scripts.csv}{name=\category,surname=\description}
%{\thecsvrow & \surname~\name & \age}

\begin{sidewaystable}
    \csvautotabular[respect all]{data/meta.csv};
    \caption{Meta variables table};
    \label{table:meta-variables};
\end{sidewaystable}

@@

\bibliographystyle{apalike}
\bibliography{refs/references-etox-base}

@@

>>>>

<WRITE ABSTRACT>

@@

@@

The \etoxbase{} consists of two parts of software: Firstly a processing pipeline \ref{fig:pipeline} downloads quarterly released new versions of the \epa{} and performs several preparation steps in R \citep{r_core_team_r_2017}, to then store the outcome in a PostgreSQL data base. This constitutes the basis upon which the R shiny web application is accessing the data \citep{r_core_team_r_2017}. The application itself is set of R functions that filter and aggregate the data according to a user's inputs \ref{fig:app}. Finally a data set, which can be downloaded and an interactive plot, for data exploration are returned. In the following the two parts are explained in detail.

@@

The data is downloaded (bd\_epa\_download.R) and then built into a PostgreSQL data base (bd\_epa\_postgres.R). Subsequently the data is imported into R for pre-processing (da\_epa1.R, da\_epa2.R, da\_epa3.R), which includes validity checks, removal of special characters, unit harmonizations and selection of relevant variables. Chemical abstract service (CAS) numbers, unique chemical identifiers together we taxon names are then used to query other data bases for additional information on chemicals and biota respectively. Chemical information is retrieved from the Pubchem data base \citep{CITE_PUBCHEM} (qu\_pc.R), from the Compendium of Pesticide Common Names \citep{CITE_AW} (qu\_aw.R), from the Physprop data base \citep{CITE_PHYSPROP} (qu\_pp.R), from EUROSTAT and the Chemspider data base \citep{CITE-CHEMSPIDER}. These queries mostly rely on the webchem R-package \citep{szocs_webchem_2015-1}. For biota, habitat and occurrence information is queried from the World Register of Marine Species (WORMS) \citep{WORMS} and from the Global Biodiversity Information Facility (GBIF) \citep{CITE_RGBIF} using the rgbif R-package \citep{chamberlain_rgbif_2018}. The acquired information is then merged (re\_merge.R), variables are combined and created (re\_combine.R), checked (re\_checks\_internal.R) and finally compiled in one table (re\_final.R) This table is then used in the application to perform search queries on. For a detailed overview of what information is collected see table \ref{tab:scripts-pipeline}.

\begin{figure}
    \includestandalone[width=\textwidth, scale=0.1]{tikz/pipeline_organigram}
    \caption{Processing Pipeline}
    \label{fig:pipeline}
\end{figure}

\subsection*{The application}
The application is accessible through \app{} and was built in R by using the shiny web application framework \citep{chang_shiny_2018}. Users can define several input parameters \ref{tab:inputs} in the application's graphical user interface (GUI), handeled in the ui.R script in order to adjust the aggregation according specific requirements. In doing so users execute R functions (fun\_filter.R, fun\_aggregation.R, fun\_filagg\_plot\_ly.R) in the server.R script, which produce a 'filtered' and an 'aggregated' data set as well as an interactive overview plot. The 'aggregated' data set is presented in the GUI. Both data sets can be downloaded via a download button.

\begin{figure}
    \includestandalone[scale=0.75]{tikz/application}
    \caption{Application}
    \label{fig:app}
\end{figure}

\subsection*{Aggregating the data}
In order to be able to calculate TUs and SSDs, derivation of single endpoint values for specific organism groups and a chemicals of the partly extensive test data, the \etoxbase{} aggregates the test results according to the chosen filters in two steps. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon plus the duration and secondly only by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the geometric mean, or the arithmetic mean as an aggregate. Besides the aggregation two cleaning can also be performed. Firstly the application allows to exclude test results exhibiting concentrations that are higher than the actual water solubility of the respective chemical at \ang{20} C. Secondly the user can choose to exclude outliers. The outliers are selected as values that exceed lower (0.25) and the upper (0.75) quartile by 1.5 times the inter-quartile range. All the steps described in this section are selectable in the GUI and are executed at each click by the server. (cf. table \ref{tab:scripts-app}).

\citep{dowle_data.table_2018}.

\subsection*{Future of the project}

@@

The code for processing the data is stored in a Github reopsitory (\git{}). All processing scripts were written in R version 3.4.4. PostgreSQL 9.5 has been used to build the data base. Hence, R together with its packages and PostgreSQL suffice to rebuild the process. The \app{} itself is stored in a second Github repostitory (\gitapp{}).
Since new versions of the \epa{} are published on a quarterly basis the data acquisition process is automated and new versions of the data and the \app{} are scheduled to be released regularly.

\begin{table}[]
    \centering
    \begin{tabular}{c|c}
         Category       & Package       & Version   & Reference \\
         processing     & RCurl         & \\
         & 
    \end{tabular}
    \caption{List of R packages used in compiling the data}
    \label{tab:rpackages}
\end{table}

@@

@@

<PUT COMPARISON WITH PPDB DATA HERE?>

\section*{Usage Notes}
%%Brief instructions that may help other researchers reuse these dataset.
%%This is an optional section, but strongly encouraged when helpful
%%to readers. This may include discussion of software packages that
%%are suitable for analyzing the assay data files, suggested downstream
%%processing steps (e.g. normalization, etc.), or tips for integrating
%%or comparing this with other datasets. If needed, authors are encouraged
%%to provide code, programs, or data processing workflows when they may help 
%%others analyse the data. We encourage authors to archive related code in 
%%a DOI-issuing archive when possible, but code may also be supplied as 
%%supplementary information files. 
%%
%%For studies involving privacy or safety controls on public access
%%to the data, this section should describe in detail these controls,
%%including how authors can apply to access the data, and what criteria
%%will be used to determine who may access the data, and any limitations
%%on data use.

The \etoxbase{} can be accessed via an application here: \app{}. The user can filter among different test parameters \ref{tab:inputs}.

The \etoxbase{} is designed to support ecotoxicologists in assessing the risk of chemicals on the environment. With an increase of ecotoxicological test data available, it becomes fundamental to provide such information in reasonable formats, meaning accessible for humans as well as machines. Currently chemical risk assessment relies on test results of a few, well tested organism groups, such as \textit{Daphnia magna}, \textit{Pimephales promelas}, \textit{Selenastrum capricornutum} reported in publications on test procedures <CITE>, compiled published data sets \citep{malaj_organic_2014} or data bases such as the Pesticide Property Data Base (PPDB) \citep{lewis_international_2016}, whereby each source having its limitations: Collecting toxicity information from individual publications is laborious and using already compiled data from publications or data bases is often limited to specific chemical groups, such as pesticides, biozides etc. The \etoxbase{}, making use of the \epa{} data aims to provide holistic aggregates (\ecfifty{}, NOEC, LOEC) of ecotoxicological tests results in adequate formats in order to support researchers to in the compilation of SSDs, TUs and other risk indicators.

Generally it has to be stated here that the returned values are subject to change, as new tests will be constantly incorporated into the data set. However, we expect values of well tested chemical-organism combinations (i.e. high N <MAYBE PUT NUMBER HERE - CITATION>) to be relatively stable. Yet, returned values from the \etoxbase{} are certainly influenced by individual tests procedures and characteristics, such as pH, temperature or conductivity to just name a few. However, using the geometric mean as an aggregate allows for an adequate estimation of the central tendency of the toxicity of a chemical towards an organism group. The geometric mean is in comparison to the arithmetic mean considerably less influenced by outliers and is suitable for skewed data. Also, the geometric mean is preferable in contrast to the median, since the median completely ignores the influence of large or small values, making it unreliable for small data sets <CITATION GEOMETRIC MEAN>.

Other initiatives, with different, though partly overlapping aims accumulating existing ecotoxicological test data are published online. The PPDB provides validated ecotoxicological test results for pesticides, narrowing chemicals to this group \citep{lewis_international_2016}. The Network of reference laboratories, research centres and related organisations for monitoring of emerging environmental substances (NORMAN) focuses on assembling river basin specific pollutants\citep{von_der_ohe_new_2011}. The EnviroTox data base \href{https://envirotoxdatabase.org/} makes also use of the \epa{} data in order to derive Predicted no-effect concentrations (PNEC) \citep{health_and_environmental_sciences_institute_hesi_envirotox_2019}. In comparison to the \etoxbase, these examples partly rely on the same data, yet not aiming to derive \ecfifty, LOEC and NOEC from it automatically.

@@

PUT SSD HERE

@@

@@

@@

%\begin{sidewaystable}
%    \csvautotabular[respect all]{data/meta.csv};
%    \caption{Meta variables table};
%    \label{table:meta-variables};
%\end{sidewaystable}

\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
Procedure   & Group       & Subgroups              & Options \\ \hline
Filter      & Compound    & CAS                    & input CAS \\ \hline
Filter      & Compound    & Concentration type     & Active Ingredient; Formulation; Total \\ \hline
Filter      & Compound    & Chemical class         & Insecticides; Herbicides; Metals \\ \hline
Filter      & Taxon       & Taxon                  & Choose a taxon \\ \hline
Filter      & Taxon       & Organism habitat       & freshwater; terrestrial; marine; brackish \\ \hline
Filter      & Taxon       & Continent              & North America; Asia; Europe \\ \hline
Filter      & Test        & Duration               & enter duration in hours \\ \hline
Filter      & Test        & Effect groups          & MOR - Mortality; POP - Population \\ \hline
Filter      & Test        & Endpoins               & EC50, LOEC, LOEL \\ \hline
Filter      & Checks      & Water solubility check & yes/no \\ \hline
Filter      & Checks      & Remove outliers        & yes/no \\ \hline
Aggregation & Aggregation & Aggregate              & Minimum; Maximum; Median \\ \hline
\end{tabular}
\caption{\etoxbase input possibilities}
\label{tab:inputs}
\end{table}

\begin{table}[]
\begin{tabular}{|l|l|}
\hline
Script                      & Description                                      \\ \hline
run_build.R                 & runs all sub scripts                             \\ \hline
setup.R                     & sets up the environment                          \\ \hline
bd\_epa\_download.R         & download EPA Ecotox data base                    \\ \hline
bd\_epa\_postgres.R         & build EPA Ecotox data base locally               \\ \hline
da\_epa1.R                  & raw EPA Ecotox data export                       \\ \hline
da\_epa2.R                  & process EPA Ecotox data                          \\ \hline
da\_epa3.R                  & reduce EPA Ecotox data                           \\ \hline
qu\_pc.R                    & query PubChem data base                          \\ \hline
qu\_aw.R                    & query Compendium of Pesticide Common Names       \\ \hline
qu\_pp.R                    & query Physprop data base                         \\ \hline
qu\_eurostat\_chem\_class.R & query Eurostat data set                          \\ \hline
qu\_chemspider\_scrape.R    & query Chemspider data base                       \\ \hline
qu\_worms2.R                & query WORMS data base                            \\ \hline
qu\_gbif.R                  & query GBIF data base                             \\ \hline
re\_merge.R                 & merge all data                                   \\ \hline
re\_combine.R               & combine information                              \\ \hline
re\_analyses.R              & ???                                              \\ \hline
re\_checks\_internal.R      & ???                                              \\ \hline
re\_final.R                 & create and write final table                     \\ \hline
wr\_dbase.R                 & write final table to data base                   \\ \hline
wr\_app.R                   & write final table and variables to app directory \\ \hline
wr\_meta.R                  & ???                                              \\ \hline
wr\_stats.R                 & ???                                              \\ \hline
\end{tabular}
\caption{Scripts that are used to process the data}
\label{tab:scripts-pipeline}
\end{table}

\begin{table}[]
\begin{tabular}{|l|l|}
\hline
Script                     & Description                                               \\ \hline
setup.R                    & sets up the environment                                   \\ \hline
ui.R                       & creates the graphical user interface of the application   \\ \hline
server.R                   & handles functions and data input and output               \\ \hline
fun\_filter.R              & function that filters the data                            \\ \hline
fun\_aggregation.R         & function that performs the aggregation                    \\ \hline
fun\_filagg\_plot\_ly.R    & creates interactive plot                                  \\ \hline
\end{tabular}
\caption{Scripts that used for the web application}
\label{tab:scripts-app}
\end{table}

@@

\bibliographystyle{apalike}
\bibliography{refs/references-etox-base,
              refs/references-etox-base-rpackages,
              refs/references-etox-base-rshinypackages}

\pagebreak
\section*{TODO}

\begin{itemize}

\item compare Etox-BAse output with PPDB

\item plot example SSD and TU

\pagebreak
\section*{MISC}

\begin{figure}
    \includestandalone[scale=0.5]{tikz/tree_source}
    \caption{Alternative processing pipeline}
    \label{fig:pipeline-tree}
\end{figure}

@@

\end{itemize}
